NURM-tool: a quantitative reproducibility measurement tool to identify numerical differences in pipelines

### 2019-04-25 paper.tex
add resuls

\title{NRM-tool (Numerical Reproducibility Measurement): 
a debugging tool to identify numerical errors 
and assess reproducibility of pipelines}

@@

\author[1]{Ali Salari}
\author[1]{Lalet Scaria}
\author[2,3]{Gregory Kiar}
\author[2]{Lindsay Lewis}
\author[1]{Tristan Glatard}
\author[2,3]{Alan C. Evans}

@@

\begin{abstract}
Many experiments show that computational analysis results are still not 
completely reproducible. In particular, computational environments 
including different operating systems, hardware categories, and 
software versions are known to have effects on the results produced by 
analysis pipelines. These effects are presumably due to the creation, 
propagation and amplification of small numerical errors across the 
pipelines. 
Although new techniques such as virtualization , version 
control systems , and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. 
However, the precise causes of such instabilities and the path along 
which they propagate in the pipelines are unclear.  We present a 
technique to identify the processes in the pipeline that create 
numerical errors along the execution, and we apply this technique to 
the HCP structural pre-processing pipelines.

@@

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. Two files will be considered numerically 
reproducible if they have identical binary contents. Binary comparison 
is calculated by comparing checksums. It must be pointed that a 
computation might be reproducible based on Peng's definition, but not 
be numerically reproducible. For instance, small numerical errors 
created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible. Results show that the 
variety of computational infrastructures including workstation types, 
parallelization methods, operating systems, and analysis packages are 
known to influence reproducibility because of the creation of small 
numerical errors~\cite{Gronenschild2012, diethelm2012limits, 
Glatard2015, bowring2018exploring}. The propagation and amplification 
of these tiny errors by analysis pipelines may cause reproducibility 
issues. 
In this case, the analysis pipelines are said to be numerically 
unstable. Numerical instability is a characteristic of the pipelines 
which amplify small numerical errors and then hamper the 
reproducibility of the analyses depending on the length of the pipeline 
and magnitude of the errors. In many cases, numerical instability is an 
important issue for reproducibility.

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. The results of comparisons indicate a significant errors 
between analysis results of the same experiment on different operating 
system. It should be noted that \emph{error} is defined as the binary 
differences between results of two analyses in this paper, which is 
calculated by the checksum method. 
The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as the main influential 
computing infrastructure on reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques to encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NRM) tool to identify the processes in 
pipeline that create numerical errors across different runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using an interposition techniques.

The reminder of this paper describe the NRM-tool in different parts as 
the way of capturing provenance information and their representation, 
the influence of different type of subjects, and how to classify 
processes and characterize errors. We describe the HCP preprocessing 
pipelines as a popular neuroimaging pipelines to test and apply the 
NRM-tool on. Finally, we report the experimental results, the processes 
that are responsible for the errors in the pipelines.

@@

We present a Numerical Reproducibility Measurement tool (NRM-tool) to 
measure numerical reproducibility of the pipelines across operating 
systems automatically (see Figure~\ref{fig:overview}). We build the 
containerized version of pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

NRM-tool produces the error JSON schema in which recognizes output files 
with errors using the checksum comparison of the results. 
We consider that a file contains errors if 
its binary comparison is different, otherwise file has no error. Also, 
we use \reprozip tool to capture the provenance information. 
Subsequently, using a iterative approach, we classify processes into 
different categories based on the state of read/write files 
with/without error. NRM-tool identifies relevant processes in the 
pipeline creating, propagating, or removing errors. We will explain 
each steps in detail in the following sections.

@@

NRM-tool reconstructs the tree of processes starting from the first process 
created by the pipeline and identifying its child processes as the ones 
that were created through \texttt{clone()} or \texttt{fork()}. It 
creates a process graph from this tree by adding edges corresponding to 
file dependencies between processes. A file dependency is defined 
between processes A and B if a file written by A is read by B. 
Figure~\ref{fig:simple_script} shows an example of a process graph 
constructed for the simple bash script illustrated in 
Algorithm~\ref{algo:sample-script} from the brain extraction procedure.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.3]{images/simple_graph}
  \caption{Process graph constructed from an example brain extraction
	script. Every node in the graph is labeled using (1) a process id 
	created by our reconstruction, (2) the name of the executable run 
	by the process. Processes that read or write temporary files are 
	represented with squares; other ones with circles. Plain edges 
	represent the process tree (\texttt{fork()} or \texttt{clone()} 
	system calls). Dashed edges represent file dependencies: temporary 
	files are in yellow and result files are in green.}
  \label{fig:simple_script}
\end{figure}

@@

Depending on the variety of computing environments, the produced 
process graph can be different. In particular, different analysis 
software package produce different process graphs for the similar 
experiments because of following different methods. We used a fixed 
analysis software package in different conditions, however, we found 
that different subjects may produce a different process graph. Among 
our dataset, we identified 4 types of subjects with different numbers 
of T1w and T2w images.

In addition, we verified that the process graphs were identical for all 
subjects of the same type in different versions of CentOS operating 
system. The remainder of the analysis is done separately for each 
subject type.

@@

NRM-tool classifies the processes into four 
categories based on the process graph and the error JSON file 
mentioned in the previous sections:
\begin{enumerate}
\item Processes that read files that do not have errors and write files 
that do not have errors are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any error but write files that have errors 
\emph{create} errors in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have errors and write files that do not have errors \emph{remove} 
errors from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have errors and write files that also have errors are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

@@

\section{Experiments}
\subsection{HCP pipelines and datasets}
\subsection{Results}

@@

>>>>

\title{NURM-tool: a quantitative reproducibility measurement tool to 
identify numerical differences in pipelines}

@@

\author[1]{Ali Salari}
\author[1]{Lalet Scaria}
\author[2,3]{Gregory Kiar}
\author[2]{Lindsay Lewis}
\author[2,3]{Alan C. Evans}
\author[1]{Tristan Glatard}

@@

\begin{abstract}
Many experiments show that computational analysis results are still not 
completely reproducible. In particular, computational environments 
including different operating systems, hardware categories, and 
software versions are known to have effects on the results produced by 
analysis pipelines. These effects are presumably due to the creation, 
propagation and amplification of small numerical differences across the 
pipelines. 
Although new techniques such as virtualization , version 
control systems , and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. 
However, the precise causes of such instabilities and the path along 
which they propagate in the pipelines are unclear.  We present a 
technique to identify the processes in the pipeline that create 
numerical differences along the execution, and we apply this technique to 
the HCP structural pre-processing pipelines.

@@

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. Two files will be considered numerically 
reproducible if they have identical binary contents. Binary comparison 
is calculated by comparing checksums. It must be pointed that a 
computation might be reproducible based on Peng's definition, but not 
be numerically reproducible. For instance, small numerical differences 
created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible. Results show that the 
variety of computational infrastructures including workstation types, 
parallelization methods, operating systems, and analysis packages are 
known to influence reproducibility because of the creation of small 
numerical differences~\cite{Gronenschild2012, diethelm2012limits, 
Glatard2015, bowring2018exploring}. The propagation and amplification 
of these slight differences by analysis pipelines may cause reproducibility 
issues. 
In this case, the analysis pipelines are said to be numerically 
unstable. Numerical instability is a characteristic of the pipelines 
which amplify small numerical differences and then hamper the 
reproducibility of the analyses depending on the length of the pipeline 
and magnitude of the differences. In many cases, numerical instability is an 
important issue for reproducibility.

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. The results of comparisons indicate a significant differences 
between analysis results of the same experiment on different operating 
system. \note{It should be noted that the binary 
differences are measured between results of two analyses in this paper, which is 
calculated by the checksum method. }
The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as the main influential 
computing infrastructure on reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques to encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using an interposition techniques.

The reminder of this paper describe the NURM-tool in different parts as 
the way of capturing provenance information and their representation, 
the influence of different type of subjects, and how to classify 
processes and characterize differences. We describe the HCP preprocessing 
pipelines as a popular neuroimaging pipelines to test and apply the 
NURM-tool on. Finally, we report the experimental results, the processes 
that are responsible for the differences in the pipelines.

@@

We present a Numerical Reproducibility Measurement tool (NURM-tool) to 
measure numerical reproducibility of the pipelines across operating 
systems automatically (see Figure~\ref{fig:overview}). We build the 
containerized version of pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

NURM-tool produces the error JSON schema in which recognizes output files 
with errors using the checksum comparison of the results. 
We consider that a file contains errors if 
its binary comparison is different, otherwise file has no error. Also, 
we use \reprozip tool to capture the provenance information. 
Subsequently, using a iterative approach, we classify processes into 
different categories based on the state of read/write files 
with/without error. NURM-tool identifies relevant processes in the 
pipeline creating, propagating, or removing errors. We will explain 
each steps in detail in the following sections.

@@

NURM-tool reconstructs the tree of processes starting from the first process 
created by the pipeline and identifying its child processes as the ones 
that were created through \texttt{clone()} or \texttt{fork()}. It 
creates a process graph from this tree by adding edges corresponding to 
file dependencies between processes. A file dependency is defined 
between processes A and B if a file written by A is read by B. 
Figure~\ref{fig:simple_script} shows an example of a process tree 
constructed for the simple bash script illustrated in 
Algorithm~\ref{algo:sample-script} from the brain extraction procedure.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.3]{images/simple_graph}
  \caption{Process tree constructed from an example brain extraction
	script. Every node in the graph is labeled using (1) a process id 
	created by our reconstruction, (2) the name of the executable run 
	by the process. Processes that read or write temporary files are 
	represented with squares; other ones with circles. Plain edges 
	represent the process tree (\texttt{fork()} or \texttt{clone()} 
	system calls). Dashed edges represent file dependencies: temporary 
	files are in yellow and result files are in green.}
  \label{fig:simple_script}
\end{figure}

@@

Depending on the variety of computing environments, different  
process tree can be produced. In particular, different analysis 
software package can construct different process trees for the 
similar experiments because of following different methods. We used a 
fixed analysis software package in different conditions, however, we 
found that different subjects may produce a different process trees.

The challenging problem is that of comparing various process trees with 
different topological structures from different subjects. To address 
this issue, we categorize subjects into different types based on the 
distances, topologically differences, between corresponding process 
trees. 
We cluster subjects into various types using the bottom-up 
hierarchical clustering approach. This approach groups together the 
subjects with similar distance, less than a threshold.

We use the tree edit distance~\cite{zhang1989simple} method to measure 
the distance between labeled trees. The distance between two labeled 
trees is defined as the minimum weighted number of edit operations that 
it takes to transform one tree to another. There are three edit 
operations that can be used to transform a tree including modify, 
remove, and insert. Modifying node \textit{n} means changing the label 
on \textit{n}. Removing a node \textit{n} means making the children of 
\textit{n} become the children of the parent of \textit{n} and then 
deleting \textit{n}. Inserting is the the complement of remove, this 
means that inserting \textit{n} as the child of \textit{n}$'$ will make 
\textit{n} the parent of a consecutive subsequence of the current 
children of \textit{n}$'$.
Each operation has an associated cost---insertions and removals each 
cost 1, and modifications cost either 1 or 2, depending on whether or 
not you consider a modification to essentially be just a removal 
followed by an insertion.

The clustering algorithm starts by finding the two subjects that are 
closest to each other on the basis of edit distance method and form the 
first cluster between these two subjects. The next step join this 
cluster to the next nearest cluster or subject which in turn results in 
another cluster. This process continues until all the subjects are 
joined together to form one big cluster or terminates when the distances 
are less than the threshold. The threshold is defined as the minimum 
distance required to separate clusters.

@@

NURM-tool classifies the processes into four 
categories based on the process graph and the error JSON file 
mentioned in the previous sections:
\begin{enumerate}
\item Processes that read files that do not have errors and write files 
that do not have errors are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any error but write files that have errors 
\emph{create} errors in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have errors and write files that do not have errors \emph{remove} 
errors from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have errors and write files that also have errors are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

@@

\section{Experiments}

\subsection{HCP pipelines and datasets}

The Human Connectome Project (HCP)~\cite{glasser2013minimal} developed 
a set of pipelines to help extraction of structural, functional or 
diffusion MRI data across a large set of high resolution MR images. 
These pre-processing pipelines create results that are available in 
standard volume and combined surface and volume spaces which makes it 
easier for researchers to compare the images across the neuroimaging 
spectrum.

We focused on HCP structural pre-processing pipelines to (1)quantify 
the effect of OS on HCP pre-processing pipelines (2) identify the 
process(es) in the pipelines that are responsible for such effects.

The structural pre-processing pipelines consists of PreFreeSurfer, 
FreeSurfer and PostFreeSurfer. In this paper, 
we analyzed PreFreeSurfer pipeline which consist of various steps 
including correction of MR 
gradient nonlinearity distortions, align the T1w and T2w images using 
FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
to T1w using FLIRT's BBR, perform a bias field correction, and register 
the subject's native structural space to MNI space.

We randomly selected 10 subjects from the HCP data release S500 which 
are described in table~\ref{table:data}.

%\begin{figure}
%\centering
%  \includegraphics[width=\columnwidth]{images/overview.png}
%  \caption{An overview of the proposed technique.}
%  \label{fig:overview}
%\end{figure}

\begin{table}
\centering
\begin{threeparttable}
\caption{An overview of the HCP data (Humman Connectome DB).}

\begin{tabular}{@{}llllll@{}}
\toprule
Subject & Release & Acquisition & Gender & Age      & Subj\_Type   \\ \midrule
103515  & Q1      & Q02         & F      & 26-30    & type1         \\
105216  & Q3      & Q03         & M      & 26-30    & type4         \\
103414  & Q2      & Q02         & F      & 22-25    & type1         \\
125525  & Q1      & Q01         & F      & 31-35    & type1         \\
142828  & Q1      & Q01         & M      & 31-35    & type1         \\
129533  & Q3      & Q04         & F      & 31-35    & type1         \\
103818  & Q1      & Q01         & F      & 31-35    & type3         \\
133928  & Q2      & Q03         & M      & 26-30    & type3         \\
148032  & Q3      & Q03         & F      & 31-35    & type2         \\
139637  & Q2      & Q03         & F      & 31-35    & type1         \\ \bottomrule
\end{tabular}
\begin{tablenotes}
      \small
      \item *Subj\_Type refers to the graph topology type of the subjects.
\end{tablenotes}
\end{threeparttable}
\label{table:data}
\end{table}

\note{Among our dataset, we identified 4 types of subjects with 
different numbers of T1w and T2w images. In addition, we verified that 
the process graphs were identical for all subjects of the same type in 
different versions of CentOS operating system. The remainder of the 
analysis is done separately for each subject type.}

\subsection{Results}

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. The Docker images were built for the HCP PreFreeSurfer 
pipeline v3.19.0 in 
CentOS 6.8 and CentOS 7.2. Docker images are available on DockerHub for 
reuse \url{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}. 
We collected the provenance trace (tree of executed processes and files 
accessed in read or write mode) for each type of 
subjects(Table~\ref{table:data}) using system-call interception as 
provided by the \reprozip tool~\cite{Chirigati2016}.

Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. An example of a pseudo-random process function is, a random 
number generator that would get initialized using a seed state. the 
proposed method can be used to identify both kind of errors. The files 
that are common to all the subjects only are taken into consideration 
for comparison. The first step is identification of files with 
differences in their checksums. This is identified using the checksums 
that are recorded after the processing. Intra-OS errors are identified 
using the run-number added as the suffix for the conditions. For 
example, the two batches of subjects processed under the same condition 
(CentOS6) are stored as run-1 and run-2. The files belonging to the 
subjects stored under the above mentioned conditions are treated as 
intra-OS runs.

\subsubsection{Inter-OS binary differences}

In the previous study~\cite{Scaria2017}, we showed that pre-processing 
pipelines of the Human Connectome Project~\cite{Glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized errors after brain 
tissue segmentation process in FreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same errors occurred in 
non-linear registration process between CentOS6 and CentOS7 in PreFreeSurfer.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized error between brain segmentation results from 
  FreeSurfer, subject 105216 (CentOS6 vs. CentOS7)~\cite{Scaria2017}.
    } 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/fnirt_result.png} 
  \caption{Absolute errors between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (CentOS6 vs. 
  CentOS7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

\subsubsection{PreFreeSurfer pipeline analysis}

We identified processes that introduce errors in PreFreeSurfer 
pipeline. Among the 117 data files produced by PreFreeSurfer, 21 did 
not have any error for any subject, 92 had errors for all subjects and 
4 had errors for 3 subjects only.

Figure \ref{fig:complete-graph} shows the annotated provenance graph 
of the PreFreeSurfer pipeline executed on CentOS6 and CentOS7. Each 
node in the graph represent an executed process in the pipeline. 
Processes that created errors are shown in red, processes that removed 
errors are in blue, and other processes are in green.  Squares denote 
processes for which the classification is uncertain, due to temporary 
files that were removed during the execution. Black edges link 
sub-processes to their parents while dashed edges denote file 
dependencies between processes (green edges: files with no errors; red 
edges: files with errors; yellow edges: temporary files).

The processes that introduce errors in PreFreeSurfer along with the 
number of occurrences are depicted in Figures~\ref{fig:pfs_table, 
fig:pfs_chart} including linear registration with “\emph{FLIRT}” (in 
ACPC Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration).  In 
addition, errors were observed in image mean and standard-deviation 
computations with “\emph{fslstats}” (in BiasFieldCorrection), and in 
masked image extrapolation with “\emph{fslmaths}” (in 
BiasFieldCorrection).  Besides, transformation format conversion with 
“\emph{convertwarp}” (in DistortionCorrection) was able to remove 
errors.

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_table.png}
  \caption{Process name along with the pipeline elements that introduce 
  errors.}
  \label{fig:pfs_table}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_chart.png} \caption{Number 
  of occurrences of errors in PreFreeSurfer on subject 103515, red and 
  blue bars indicate the processes which create and remove errors 
  respectively.} 
  \label{fig:pfs_chart}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=1\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreesurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure}
